UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-03 ms/main.tex
ciiiites

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds. UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.

@@

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Colouring the projection by distance from London reflects the high migration levels to the city and surrounding area; individuals with more admixed backgrounds tend to have coordinates in the south east of the UK, as visible in figure~\ref{fig:supp_london_distance}. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.

>>>>

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds. Comparing UMAP to PCA with this data, the former captures local and fine-scale structure well while the latter identifies major axes of variation in figure~\ref{fig:fig_ukbb}. UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.

@@

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe and British-wide data\cite{novembre2008europe,leslie2015fine}. Colouring the projection by distance from London reflects the high migration levels to the city and surrounding area; individuals with more admixed backgrounds tend to have coordinates in the south east of the UK, as visible in figure~\ref{fig:supp_london_distance}. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.
